Parsing Postfix log files

### 2008-03-05 doc/logparser.tex
Corrections after proofreading section 8.

@@

Parsing Postfix logs appears at first sight to be an almost trivial task,
especially given previous experience in parsing logs, but it turns out to
be a much more taxing project than initially expected.  The variety and
breadth of log lines produced by Postfix is quite surprising, as a quick
survey of sample log files gives the impression that the number of distinct
log lines is quite small; this impression is due to the uneven distribution
exhibited by logs produced in normal operation (see graph~\refwithpage{rule
hits graph} for a vivid illustration of this).

Given the diverse nature of the log lines and the ease with which
administrators can cause new lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Despite the
resulting initial increase in complexity the job is quite tractable, though
it does raise interesting efficiency and optimisation questions (answered
in \sectionref{rule efficiency}).  Providing a tool (\sectionref{creating
new rules}) to ease the generation of \regexes{} from unparsed log lines
should greatly help users add rules to parse formerly unparsed lines.

The implementation not only substantially eases the parsing of new log
lines, it makes adding new actions (\sectionref{parsing-algorithm}) an
almost trivial task.  The simplicity of adding a new action
(\sectionref{adding new actions}) frees the implementor from worrying about
how their action might disrupt parsing of other lines or the behaviour of
other actions, allowing their concentration to be focused on correctly
implementing their action.

The division of the parser into rules and actions is unusual because it
divides the two so completely; although parsers are often divided (lex and
yacc~\cite{lex-and-yacc} based parsers being an obvious example), the parts
are usually quite internally interdependent, and will be combined into a
complete parser by the compilation process; in contrast \parsername{} keeps
the rules and actions separate until the parser runs.  XXX EXTEND THIS\@;
SOMETHING ABOUT THE CONTROL FLOW BEING DIFFERENT, I.E. RULE DETERMINES
ACTION, RATHER THAN ACTION EXPECTING INPUT\@.

The emergent behaviour~\cite{Wikipedia-Emergence} exhibited by the rules
and actions is also interesting, and is discussed beside the flow chart
depicting said emergent behaviour (\sectionref{flow-chart}).  This emergent
behaviour greatly eases

The real difficulties arise once the parser is successfully dealing with
90\% of the log lines, as the irregularities and complications previously
explained only begin to become apparent once the vast majority of log lines
have been parsed successfully.  Dealing with numerous, infrequently
occurring log lines is a simple task, albeit tiresome, whereas dealing with
mails where information appears to be missing is much more grueling.
Trawling through the logs, looking for something out of the ordinary,
possibly hundreds or even thousands of lines away from the last mention of
the queueid in question, is extremely time consuming and error prone.  In
many cases the task is not to spot the line which is unusual, but to spot
that a line normally present is missing.  In all cases the evidence must be
used to construct a hypothesis to explain the irregularities, and that
hypothesis must then be tested in the parser; if successful the parser must
be modified to deal with the irregularities, without adversely affecting
the existing functionality.  The complications described in
\sectionref{additional complications} were solved in the order they are
described in, and that order closely resembles the frequency with which
they occur; the most frequently occurring complications dominate the
warning messages produced, and so naturally they are the first
complications to be dealt with.

>>>>

\label{Emergent behaviour}

@@

Parsing Postfix logs appears at first sight to be an almost trivial task,
especially if one has previous experience in parsing logs, but it turns out
to be a much more taxing project than initially expected.  The variety and
breadth of log lines produced by Postfix is quite surprising, because a
quick survey of sample log files gives the impression that the number of
distinct log lines is quite small; this impression is due to the uneven
distribution exhibited by logs produced in normal operation (see
graph~\refwithpage{rule hits graph} for a vivid illustration of this).

Given the diverse nature of the log lines and the ease with which
administrators can cause new lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Despite the
resulting initial increase in complexity the task is quite tractable,
though it does raise interesting efficiency and optimisation questions
(answered in \sectionref{rule efficiency}).  Providing a tool
(\sectionref{creating new rules}) to ease the generation of \regexes{} from
unparsed log lines should greatly help users add rules to parse formerly
unparsed log lines.

The implementation not only substantially eases the parsing of new log
lines, it makes adding new actions (\sectionref{adding new actions}) a
relatively easy task.  The simplicity of adding a new action frees the
implementor from worrying about how their action might disrupt parsing of
other log lines or the behaviour of other actions, allowing their
concentration to focus on correctly implementing their new action.

The division of the parser into rules, actions and framework is unusual
because rules are separated so completely from the actions and framework;
although parsers are often divided (lex and yacc~\cite{lex-and-yacc} based
parsers being an obvious example), the parts are usually quite internally
interdependent, and will be combined into a complete parser by the
compilation process; in contrast \parsername{} keeps the rules and actions
separate until the parser runs.  The actions and framework are not as
completely separated, as the actions depend on services provided by the
framework; however the actions and framework are not tightly integrated: it
would be possible, with some work, to completely separate the two.

XXX EXTEND THIS\@; SOMETHING ABOUT THE CONTROL FLOW BEING UNUSUAL, I.E.\
RULE DETERMINES ACTION, RATHER THAN ACTION EXPECTING INPUT\@.

The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
behaviour}) exhibited by the rules and actions is also interesting, and is
discussed after the flow chart (\sectionref{flow-chart}) and explanation of
the paths through the parser.  This emergent behaviour greatly eases adding
new actions (\sectionref{adding new actions}), as the actions do not have
to be inserted into an explicit flow of control; new actions will naturally
find their place in the paths through the parser available to a mail.

The real difficulties arise once the parser is successfully dealing with
90\% of the log lines, as the irregularities and complications previously
explained only begin to become apparent once the vast majority of log lines
have been parsed successfully.  Adding new rules to deal with numerous,
infrequently occurring log lines is a simple task, albeit tiresome (though
alleviated by the tool described in \sectionref{adding new actions}),
whereas dealing with mails where information appears to be missing is much
more grueling.  Trawling through the logs, looking for something out of the
ordinary, possibly hundreds or even thousands of lines away from the last
mention of the queueid in question, is extremely time consuming and error
prone.  In some cases the task is not to spot the line which is unusual,
but to spot that a line normally present is missing, i.e.\ to realise that
one line amongst hundreds is absent.  In all cases the evidence must be
used to construct a hypothesis to explain the irregularities, and that
hypothesis must then be tested in the parser; if successful the parser must
be modified to deal with the irregularities, without adversely affecting
the existing functionality.  The complications described in
\sectionref{additional complications} were solved in the order they are
described in, and that order closely resembles the frequency with which
they occur; the most frequently occurring complications dominate the
warning messages produced, and so naturally they are the first
complications to be dealt with.

A parser's ability to correctly parse it's input is extremely important;
the parser's coverage of its test log files is discussed in
\sectionref{parsing coverage}.  Both its success at parsing individual log
lines (\sectionref{log-lines-covered}) and its correctness in
reconstructing each mail's journey through Postfix
(\sectionref{mails-covered}) are described in detail, including the results
of manually verifying the correct parsing of a subset of the test log
files.

\newpage
